---
title: "Finding America's Most Average Counties: A Philosophical Investigation of Demographic Identity"
subtitle: "The Goldilocks Zone of American Ordinariness"
author: "Census Data Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
                      fig.width = 12, fig.height = 8)

# Load required libraries
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(viridis)
library(knitr)
library(kableExtra)
library(gt)
library(ggrepel)
library(patchwork)
library(plotly)
library(scales)
library(corrplot)
library(ggradar)
library(ggforce)
library(DT)
```

# The Quest for Perfect Ordinariness

> *"What is the most American place in America? Where do we find the statistical soul of our nation?"*

In an era of increasing polarization and geographic sorting, this analysis embarks on a whimsical yet profound journey to answer a deceptively simple question: **Which American counties are the most perfectly, comprehensively average?**

This is not merely a statistical exercise. It's a philosophical investigation into the nature of American identity itself. In a country where coastal elites and rural heartlanders seem to inhabit different realities, where red and blue states appear to diverge into separate nations, we search for the places that remain—statistically speaking—quintessentially representative of us all.

**The Central Paradox**: In seeking the most ordinary places, we may discover something extraordinary about what it means to be American in the 21st century.

## The Methodological Journey: Wrestling with "Average"

What does it mean to be average? This seemingly simple question reveals layers of complexity that illuminate the deeper challenge of defining American identity.

### The Evolution of Our Approach

Our investigation employs multiple lenses to capture different dimensions of "averageness":

1. **Mahalanobis Distance**: Accounts for correlations between variables—a county might be average on education and income individually, but their combination might be unusual
2. **Euclidean Distance**: Simple geometric distance from national means
3. **Median Absolute Deviation**: Robust to outliers, focuses on typical deviations
4. **Quartile Spread**: Measures consistency across the distribution of differences
5. **Maximum Deviation**: Identifies counties with no extreme outliers

Each metric tells a different story about what makes a place "typical," revealing that averageness itself is multifaceted.

## Executive Summary

**Key Discoveries:**

- **Van Buren County, Michigan** emerges as America's most comprehensively average county across 40+ demographic dimensions
- **Perfect averageness is rare**: Only a handful of counties achieve comprehensive demographic balance
- **Geographic clustering**: The most average counties concentrate in the Midwest and parts of the South
- **Temporal shifts**: Places that were average in 2010 are not necessarily average today—the definition of "typical America" is evolving
- **The averageness paradox**: Being thoroughly ordinary requires an extraordinary balancing act across dozens of variables

```{r load-data, include=FALSE}
# Try to load enhanced data, fallback to original
enhanced_file <- "/Users/dmitryshkolnik/Projects/census-monkey-typewriter/analyses/whimsical/goldilocks-zone/data/goldilocks_analysis_enhanced.RData"
original_file <- "/Users/dmitryshkolnik/Projects/census-monkey-typewriter/analyses/whimsical/goldilocks-zone/data/goldilocks_analysis.RData"

if (file.exists(enhanced_file)) {
  load(enhanced_file)
} else if (file.exists(original_file)) {
  load(original_file)
  cat("Using original analysis data - enhanced features may not be available\\n")
} else {
  stop("Neither enhanced nor original analysis data found - please run the R script first")
}

# Define consistent visual theme
theme_goldilocks <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", margin = margin(b = 20)),
    plot.subtitle = element_text(size = 12, color = "grey40", margin = margin(b = 15)),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    plot.margin = margin(15, 15, 15, 15),
    strip.text = element_text(face = "bold", size = 11)
  )

# Load county geometries for mapping
us_states <- tigris::states(cb = TRUE, resolution = "20m") %>%
  shift_geometry() %>%
  filter(!STUSPS %in% c("AK", "HI", "PR", "VI", "AS", "GU", "MP"))
```

# The Discovery Process: How Different Definitions Change Everything

## Multiple Metrics, Multiple Stories

```{r metrics-comparison, fig.height=10}
# Create comparison of different averageness metrics
metrics_data <- county_analysis %>%
  left_join(
    county_data %>% 
      select(GEOID, NAME) %>% 
      st_drop_geometry(), 
    by = "GEOID"
  ) %>%
  select(NAME, mahal_distance, euclidean_distance, mad_score, quartile_score, max_deviation) %>%
  filter(!is.na(NAME))

# Get top 10 by each metric
top_by_metric <- list(
  "Mahalanobis\n(Correlation-Adjusted)" = metrics_data %>% arrange(mahal_distance) %>% slice_head(n = 10),
  "Euclidean\n(Simple Distance)" = metrics_data %>% arrange(euclidean_distance) %>% slice_head(n = 10),
  "Median Deviation\n(Robust)" = metrics_data %>% arrange(mad_score) %>% slice_head(n = 10),
  "Quartile Spread\n(Consistency)" = metrics_data %>% arrange(quartile_score) %>% slice_head(n = 10),
  "Maximum Deviation\n(No Outliers)" = metrics_data %>% arrange(max_deviation) %>% slice_head(n = 10)
)

# Create combined plot showing top counties by each metric
metric_plots <- map2(top_by_metric, names(top_by_metric), ~{
  .x %>%
    mutate(
      rank = row_number(),
      county_short = str_extract(NAME, "^[^,]+")
    ) %>%
    ggplot(aes(x = reorder(county_short, -rank), y = rank)) +
    geom_col(fill = "grey20", alpha = 0.7) +
    geom_text(aes(label = rank), hjust = -0.1, color = "white", fontface = "bold") +
    coord_flip() +
    labs(
      title = .y,
      x = "",
      y = "Rank"
    ) +
    theme_goldilocks +
    theme(
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 12),
      axis.title.x = element_text(size = 10)
    )
})

# Combine plots
wrap_plots(metric_plots, ncol = 3, nrow = 2) +
  plot_annotation(
    title = "The Many Faces of Averageness: Top Counties by Different Metrics",
    subtitle = "Each definition of 'average' reveals different winners, exposing the complexity of typicality",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "grey40")
    )
  )
```

**The Revelation**: Different mathematical definitions of "average" produce dramatically different results. This isn't a flaw—it's a feature that reveals the multidimensional nature of demographic typicality.

- **Mahalanobis distance** rewards counties that balance correlated variables appropriately
- **Euclidean distance** treats all variables as equally important
- **Median deviation** focuses on consistency across the middle of distributions
- **Quartile spread** identifies counties with narrow ranges of deviations
- **Maximum deviation** finds places with no extreme outliers

## The Consistency Test: Counties That Appear Average by Any Measure

```{r consistency-analysis}
# Analyze which counties appear consistently across metrics
all_top_counties <- unlist(map(top_by_metric, ~.x$NAME))
consistency_table <- table(all_top_counties)
consistent_counties <- sort(consistency_table[consistency_table > 1], decreasing = TRUE)

consistent_df <- data.frame(
  County = names(consistent_counties),
  Appearances = as.numeric(consistent_counties),
  Percentage = round(as.numeric(consistent_counties) / 5 * 100, 1)
) %>%
  arrange(desc(Appearances))

consistent_df %>%
  gt() %>%
  tab_header(
    title = "The Consistently Average: Counties Appearing in Multiple Top-10 Lists",
    subtitle = "These places achieve averageness by any mathematical definition"
  ) %>%
  cols_label(
    County = "County",
    Appearances = "Times in Top 10",
    Percentage = "% of Metrics"
  ) %>%
  tab_style(
    style = cell_fill(color = "#e8f4f8"),
    locations = cells_body(rows = Appearances >= 3)
  ) %>%
  tab_source_note("Source: Analysis of 5 different averageness metrics") %>%
  fmt_percent(columns = Percentage, scale_values = FALSE)
```

**The True Champions of Ordinariness**: Counties appearing in multiple top-10 lists represent places that achieve averageness by any reasonable definition—the statistical equivalent of being fluent in multiple languages of typicality.

# The Geographic Soul of America

## Mapping the Goldilocks Zone

```{r averageness-map, fig.height=10}
# Prepare county data for mapping
county_map_data <- county_data %>%
  left_join(
    county_analysis %>% 
      select(GEOID, mahal_distance), 
    by = "GEOID"
  ) %>%
  filter(!is.na(mahal_distance))

# Create the main averageness map
main_map <- ggplot() +
  geom_sf(data = us_states, fill = "white", color = "grey90", size = 0.3) +
  geom_sf(data = county_map_data, 
          aes(fill = mahal_distance), 
          color = "white", size = 0.02) +
  scale_fill_viridis_c(
    name = "Averageness\nScore",
    option = "plasma",
    trans = "log10",
    labels = function(x) sprintf("%.1f", x),
    guide = guide_colorbar(
      title.position = "top",
      barwidth = 1,
      barheight = 12
    )
  ) +
  labs(
    title = "America's Geographic Distribution of Demographic Averageness",
    subtitle = "The darker the county, the closer it is to perfect demographic ordinariness"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, color = "grey40", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )

# Add inset showing top 20 counties
top_20_counties <- county_analysis %>%
  arrange(mahal_distance) %>%
  slice_head(n = 20) %>%
  left_join(
    county_data %>% 
      select(GEOID, NAME) %>% 
      st_drop_geometry(), 
    by = "GEOID"
  ) %>%
  left_join(
    county_data %>% select(GEOID, geometry),
    by = "GEOID"
  ) %>%
  st_sf()

# Create focused map of most average counties
inset_map <- ggplot() +
  geom_sf(data = us_states, fill = "grey95", color = "grey80", size = 0.2) +
  geom_sf(data = top_20_counties, 
          aes(fill = mahal_distance), 
          color = "red", size = 0.8) +
  scale_fill_viridis_c(
    option = "plasma",
    guide = "none"
  ) +
  labs(
    title = "Top 20 Most Average Counties"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
    plot.margin = margin(5, 5, 5, 5)
  )

# Combine maps
main_map + 
  inset_element(inset_map, left = 0.02, bottom = 0.52, right = 0.35, top = 0.98)
```

**Geographic Patterns of Ordinariness:**

1. **The Midwest Dominance**: Michigan, Ohio, Wisconsin, and Iowa emerge as the heartland of American averageness
2. **Southern Balance**: The Carolinas and parts of the South contribute significantly to the Goldilocks Zone
3. **Coastal Extremes**: East and West Coast counties tend toward demographic distinctiveness
4. **Western Variability**: Mountain West shows high variance, with few comprehensively average places

## Regional Clustering of Averageness

```{r regional-analysis, fig.height=8}
# Analyze averageness by region
county_regional <- county_analysis %>%
  left_join(
    county_data %>% 
      select(GEOID, NAME) %>% 
      st_drop_geometry(), 
    by = "GEOID"
  ) %>%
  mutate(
    state = str_extract(NAME, "[A-Z]{2}$"),
    region = case_when(
      state %in% c("ME", "NH", "VT", "MA", "RI", "CT", "NY", "NJ", "PA") ~ "Northeast",
      state %in% c("OH", "MI", "IN", "WI", "IL", "MN", "IA", "MO", "ND", "SD", "NE", "KS") ~ "Midwest",
      state %in% c("DE", "MD", "DC", "VA", "WV", "KY", "TN", "NC", "SC", "GA", "FL", "AL", "MS", "AR", "LA", "OK", "TX") ~ "South",
      state %in% c("MT", "WY", "CO", "NM", "ID", "UT", "NV", "AZ", "WA", "OR", "CA", "AK", "HI") ~ "West",
      TRUE ~ "Other"
    )
  ) %>%
  filter(region != "Other")

# Regional distribution plot
regional_plot <- county_regional %>%
  ggplot(aes(x = region, y = mahal_distance)) +
  geom_violin(aes(fill = region), alpha = 0.7, color = "grey20") +
  geom_boxplot(width = 0.3, alpha = 0.8, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("#2c7fb8", "#41b6c4", "#7fcdbb", "#c7e9b4")) +
  scale_y_log10(labels = number_format(accuracy = 1)) +
  labs(
    title = "Regional Distribution of Demographic Averageness",
    subtitle = "Midwest counties cluster toward greater averageness, while coastal regions show more variation",
    x = "Region",
    y = "Averageness Score (log scale)",
    caption = "Lower scores indicate greater averageness"
  ) +
  theme_goldilocks +
  theme(legend.position = "none")

# Top counties by region
regional_champions <- county_regional %>%
  group_by(region) %>%
  arrange(mahal_distance) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  select(region, NAME, mahal_distance) %>%
  mutate(
    rank_in_region = rep(1:3, 4),
    county_short = str_extract(NAME, "^[^,]+")
  )

regional_table <- regional_champions %>%
  select(region, rank_in_region, county_short, mahal_distance) %>%
  pivot_wider(names_from = rank_in_region, values_from = c(county_short, mahal_distance), names_sep = "_") %>%
  select(region, county_short_1, mahal_distance_1, county_short_2, mahal_distance_2, county_short_3, mahal_distance_3)

regional_table %>%
  gt() %>%
  tab_header(
    title = "Regional Champions of Averageness",
    subtitle = "Most average counties within each major U.S. region"
  ) %>%
  cols_label(
    region = "Region",
    county_short_1 = "1st",
    mahal_distance_1 = "Score",
    county_short_2 = "2nd", 
    mahal_distance_2 = "Score",
    county_short_3 = "3rd",
    mahal_distance_3 = "Score"
  ) %>%
  fmt_number(columns = starts_with("mahal_distance"), decimals = 2) %>%
  tab_spanner(label = "Most Average", columns = c(county_short_1, mahal_distance_1)) %>%
  tab_spanner(label = "Second Most Average", columns = c(county_short_2, mahal_distance_2)) %>%
  tab_spanner(label = "Third Most Average", columns = c(county_short_3, mahal_distance_3))

regional_plot
```

# The Evolution of Averageness: 2010 vs 2022

## How America's Center Has Shifted

```{r temporal-evolution, fig.height=10}
# Compare 2010 vs 2022 top counties
temporal_comparison_plot <- temporal_comparison %>%
  filter(!is.na(NAME)) %>%
  ggplot(aes(x = rank_2010, y = rank_2022)) +
  geom_point(alpha = 0.6, size = 1.5, color = "grey20") +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "solid", alpha = 0.7) +
  geom_text_repel(
    data = temporal_comparison %>% 
      filter(!is.na(NAME) & (rank_2010 <= 50 | rank_2022 <= 50)) %>%
      filter(abs(rank_change) > 500),
    aes(label = str_extract(NAME, "^[^,]+")),
    size = 3,
    max.overlaps = 10
  ) +
  scale_x_continuous(labels = comma_format()) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "The Shifting Geography of American Averageness (2010 vs 2022)",
    subtitle = "Counties below the blue line became more average; those above became less average",
    x = "2010 Averageness Rank",
    y = "2022 Averageness Rank",
    caption = "Each point represents a county. Red line shows correlation; blue line shows no change."
  ) +
  theme_goldilocks

# Biggest changes plot
changes_plot <- bind_rows(
  biggest_improvers %>% mutate(direction = "Became More Average"),
  biggest_decliners %>% mutate(direction = "Became Less Average")
) %>%
  mutate(
    county_short = str_extract(NAME, "^[^,]+"),
    abs_change = abs(rank_change)
  ) %>%
  ggplot(aes(x = reorder(county_short, abs_change), y = abs_change, fill = direction)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Became More Average" = "#2c7fb8", "Became Less Average" = "#d73027")) +
  coord_flip() +
  labs(
    title = "Biggest Shifts in Averageness (2010-2022)",
    subtitle = "Counties with the largest changes in their demographic typicality",
    x = "",
    y = "Change in Rank (absolute value)",
    fill = "Direction"
  ) +
  theme_goldilocks +
  theme(legend.position = "bottom")

# Combine temporal plots
temporal_comparison_plot / changes_plot +
  plot_layout(heights = c(2, 1))
```

**Key Temporal Insights:**

1. **Demographic stability**: The correlation between 2010 and 2022 rankings suggests some consistency in what makes places average
2. **Notable shifts**: Some counties have dramatically moved toward or away from averageness as America's demographic center has evolved
3. **Regional patterns**: Changes often reflect broader demographic trends like urbanization, aging, and economic transformation

## What Changed About "Average"?

```{r national-means-comparison, fig.height=6}
# Compare national means between 2010 and 2022
means_comparison <- bind_rows(
  national_means_2010 %>% 
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    mutate(year = 2010),
  national_means %>%
    select(names(national_means_2010)) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    mutate(year = 2022)
) %>%
  mutate(
    variable_label = case_when(
      variable == "pct_white" ~ "% White",
      variable == "pct_black" ~ "% Black",
      variable == "pct_hispanic" ~ "% Hispanic",
      variable == "median_income" ~ "Median Income",
      variable == "unemployment_rate" ~ "Unemployment Rate",
      variable == "poverty_rate" ~ "Poverty Rate",
      variable == "pct_bachelors" ~ "% Bachelor's Degree",
      variable == "pct_owner_occupied" ~ "% Homeownership",
      TRUE ~ variable
    )
  ) %>%
  filter(variable != "pct_male")  # Remove for clarity

means_comparison %>%
  ggplot(aes(x = year, y = value, color = variable_label, group = variable_label)) +
  geom_line(size = 2, alpha = 0.8) +
  geom_point(size = 3) +
  facet_wrap(~variable_label, scales = "free_y", ncol = 4) +
  scale_color_viridis_d(option = "plasma", guide = "none") +
  scale_x_continuous(breaks = c(2010, 2022)) +
  labs(
    title = "How America's Demographic Center Has Shifted (2010-2022)",
    subtitle = "Changes in national means reveal evolving definition of 'average American'",
    x = "Year",
    y = "Value",
    caption = "Each panel shows how the national average changed for that variable"
  ) +
  theme_goldilocks +
  theme(
    strip.text = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )
```

# Deep Dive: Profiles of America's Most Average Places

## Virtual Tours of the Goldilocks Zone

```{r county-profiles, results='asis'}
# Create detailed profiles for top 5 counties
top_5_detailed <- county_full_analysis %>%
  arrange(mahal_distance) %>%
  slice_head(n = 5)

for (i in 1:5) {
  county <- top_5_detailed[i, ]
  
  cat("\n\n## Profile #", i, ": ", county$NAME, "\n\n")
  cat("**Averageness Score:**", round(county$mahal_distance, 3), "\n\n")
  
  # Create radar chart data
  radar_vars <- c("pct_white", "pct_black", "pct_hispanic", "median_income", 
                  "pct_bachelors", "unemployment_rate", "poverty_rate", "pct_owner_occupied")
  
  county_radar_data <- county %>%
    select(all_of(radar_vars)) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "county_value") %>%
    left_join(
      national_means %>%
        select(all_of(radar_vars)) %>%
        pivot_longer(everything(), names_to = "variable", values_to = "national_value"),
      by = "variable"
    ) %>%
    mutate(
      # Normalize to 0-100 scale for radar chart
      county_norm = case_when(
        variable == "median_income" ~ county_value / 100000 * 100,
        TRUE ~ county_value
      ),
      national_norm = case_when(
        variable == "median_income" ~ national_value / 100000 * 100,
        TRUE ~ national_value
      ),
      variable_label = case_when(
        variable == "pct_white" ~ "% White",
        variable == "pct_black" ~ "% Black",
        variable == "pct_hispanic" ~ "% Hispanic",
        variable == "median_income" ~ "Median Income\n(scaled)",
        variable == "pct_bachelors" ~ "% Bachelor's",
        variable == "unemployment_rate" ~ "Unemployment",
        variable == "poverty_rate" ~ "Poverty Rate",
        variable == "pct_owner_occupied" ~ "% Homeownership"
      )
    )
  
  # What makes this county special
  closest_matches <- county_radar_data %>%
    mutate(abs_diff = abs(county_value - national_value)) %>%
    arrange(abs_diff) %>%
    slice_head(n = 3)
  
  cat("**Closest Matches to National Average:**\n")
  for (j in 1:nrow(closest_matches)) {
    match <- closest_matches[j, ]
    cat("-", match$variable_label, ": ", 
        sprintf("%.1f", match$county_value), 
        " (national: ", sprintf("%.1f", match$national_value), 
        ", difference: ", sprintf("%+.1f", match$county_value - match$national_value), ")\n")
  }
  
  cat("\n")
}
```

## The Radar Chart Revelation

```{r radar-charts, fig.height=12}
# Create radar charts for top 5 counties
create_radar_data <- function(county_data, county_name) {
  radar_vars <- c("pct_white", "pct_black", "pct_hispanic", "median_income", 
                  "pct_bachelors", "unemployment_rate", "poverty_rate", "pct_owner_occupied")
  
  county_values <- county_data %>%
    select(all_of(radar_vars)) %>%
    as.numeric()
  
  national_values <- national_means %>%
    select(all_of(radar_vars)) %>%
    as.numeric()
  
  # Normalize income to percentage scale
  county_values[4] <- county_values[4] / 1000  # Convert to thousands
  national_values[4] <- national_values[4] / 1000
  
  data.frame(
    County = county_values,
    National = national_values,
    variable = c("% White", "% Black", "% Hispanic", "Income (1000s)", 
                 "% Bachelor's", "Unemployment", "Poverty", "% Own Home")
  ) %>%
    column_to_rownames("variable") %>%
    t() %>%
    as.data.frame()
}

# Create individual radar plots for top 5
radar_plots <- map(1:5, ~{
  county <- top_5_detailed[.x, ]
  radar_data <- create_radar_data(county, county$NAME)
  
  # Add min/max rows for scaling
  radar_data_scaled <- rbind(
    max = c(100, 50, 50, 100, 50, 20, 30, 100),  # Reasonable maxima
    min = c(0, 0, 0, 0, 0, 0, 0, 0),             # Minima
    radar_data
  )
  
  # Create the plot using base plotting (ggradar alternative)
  par(mfrow = c(1, 1), mar = c(1, 1, 3, 1))
  
  # Simple visualization alternative
  county_name <- str_extract(county$NAME, "^[^,]+")
  
  # Create a data frame for ggplot radar alternative
  radar_long <- radar_data %>%
    rownames_to_column("Type") %>%
    pivot_longer(-Type, names_to = "Variable", values_to = "Value") %>%
    mutate(
      Variable = factor(Variable, levels = unique(Variable)),
      Type = factor(Type, levels = c("County", "National"))
    )
  
  ggplot(radar_long, aes(x = Variable, y = Value, color = Type, group = Type)) +
    geom_polygon(aes(fill = Type), alpha = 0.3) +
    geom_point(size = 3) +
    geom_line(size = 1.2) +
    scale_color_manual(values = c("County" = "#2c7fb8", "National" = "#d73027")) +
    scale_fill_manual(values = c("County" = "#2c7fb8", "National" = "#d73027")) +
    coord_polar() +
    labs(
      title = paste("Profile:", county_name),
      subtitle = paste("Averageness Score:", round(county$mahal_distance, 3))
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 9),
      axis.text.y = element_blank(),
      axis.title = element_blank(),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5),
      legend.position = "bottom",
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank()
    )
})

# Arrange radar plots
wrap_plots(radar_plots, ncol = 3, nrow = 2) +
  plot_annotation(
    title = "Radar Profiles of America's Most Average Counties",
    subtitle = "Blue shows county values, red shows national averages - closer overlap indicates greater averageness",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "grey40")
    )
  )
```

# The Extremes: What Makes Places Least Average?

## Counter-Examples: America's Most Distinctive Counties

```{r extreme-counties-analysis, fig.height=10}
# Analyze the most extreme counties
extreme_top_10 <- extreme_counties %>%
  slice_head(n = 10) %>%
  mutate(county_short = str_extract(NAME, "^[^,]+"))

# What makes them extreme?
extreme_deviations <- county_analysis %>%
  filter(GEOID %in% extreme_top_10$GEOID) %>%
  left_join(
    county_data %>% 
      select(GEOID, NAME) %>% 
      st_drop_geometry(), 
    by = "GEOID"
  ) %>%
  select(-euclidean_distance, -mad_score, -quartile_score, -max_deviation) %>%
  pivot_longer(-c(GEOID, NAME, mahal_distance), names_to = "variable", values_to = "value") %>%
  left_join(
    national_means %>%
      pivot_longer(everything(), names_to = "variable", values_to = "national_mean"),
    by = "variable"
  ) %>%
  mutate(
    deviation = abs(value - national_mean),
    standardized_deviation = deviation / national_mean,
    county_short = str_extract(NAME, "^[^,]+")
  ) %>%
  group_by(GEOID, county_short) %>%
  arrange(desc(standardized_deviation)) %>%
  slice_head(n = 3) %>%
  ungroup()

# Extreme counties table
extreme_counties %>%
  slice_head(n = 10) %>%
  select(NAME, mahal_distance) %>%
  mutate(
    rank = row_number(),
    county_short = str_extract(NAME, "^[^,]+")
  ) %>%
  gt() %>%
  tab_header(
    title = "America's Most Distinctive Counties",
    subtitle = "Places that deviate most dramatically from national demographic norms"
  ) %>%
  cols_label(
    rank = "Rank",
    NAME = "County",
    mahal_distance = "Extremeness Score"
  ) %>%
  fmt_number(columns = mahal_distance, decimals = 1) %>%
  tab_style(
    style = cell_fill(color = "#fee8c8"),
    locations = cells_body(rows = 1:5)
  ) %>%
  tab_footnote(
    footnote = "Higher scores indicate greater deviation from national averages",
    locations = cells_column_labels(columns = mahal_distance)
  )

# Visualization of extremes
extreme_viz <- extreme_deviations %>%
  ggplot(aes(x = reorder(county_short, -standardized_deviation), 
             y = standardized_deviation, 
             fill = variable)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_viridis_d(option = "plasma", name = "Variable") +
  coord_flip() +
  labs(
    title = "What Makes Counties Extreme?",
    subtitle = "Biggest deviations from national averages for most distinctive places",
    x = "County",
    y = "Standardized Deviation from National Mean",
    caption = "Each county shows its top 3 most extreme variables"
  ) +
  theme_goldilocks +
  theme(legend.position = "bottom")

extreme_viz
```

# Variable Sensitivity: The Fragility of Averageness

## How Rankings Change with Different Variable Sets

```{r sensitivity-analysis, fig.height=8}
# Show how different variable combinations affect rankings
sensitivity_data <- map_dfr(names(sensitivity_results), ~{
  data.frame(
    variable_set = .x,
    top_counties = sensitivity_results[[.x]][1:5],  # Top 5 for each set
    rank = 1:5
  )
}) %>%
  left_join(
    county_data %>% 
      select(GEOID, NAME) %>% 
      st_drop_geometry(), 
    by = c("top_counties" = "GEOID")
  ) %>%
  mutate(
    county_short = str_extract(NAME, "^[^,]+"),
    variable_set_label = case_when(
      variable_set == "demographics_only" ~ "Demographics\nOnly",
      variable_set == "economics_only" ~ "Economics\nOnly", 
      variable_set == "no_race" ~ "No Race\nVariables",
      variable_set == "no_income" ~ "No Income\nVariables",
      variable_set == "core_only" ~ "Core Variables\nOnly"
    )
  )

# Heatmap of county appearances across variable sets
sensitivity_heatmap <- sensitivity_data %>%
  select(variable_set_label, rank, county_short) %>%
  pivot_wider(names_from = variable_set_label, values_from = rank, values_fill = NA) %>%
  column_to_rownames("county_short") %>%
  as.matrix()

# Convert to long format for ggplot
heatmap_long <- sensitivity_heatmap %>%
  as.data.frame() %>%
  rownames_to_column("County") %>%
  pivot_longer(-County, names_to = "Variable_Set", values_to = "Rank") %>%
  filter(!is.na(Rank))

heatmap_plot <- ggplot(heatmap_long, aes(x = Variable_Set, y = County, fill = Rank)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = Rank), color = "white", fontface = "bold") +
  scale_fill_viridis_c(option = "plasma", direction = -1, name = "Rank") +
  labs(
    title = "Sensitivity Analysis: How Variable Selection Changes Rankings",
    subtitle = "Counties appearing in top 5 across different variable combinations",
    x = "Variable Set",
    y = "County"
  ) +
  theme_goldilocks +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

heatmap_plot
```

**The Fragility Insight**: Rankings are highly sensitive to variable selection, revealing that "averageness" depends critically on what we choose to measure. This sensitivity analysis exposes a fundamental question: **What variables truly define American identity?**

# The Philosophy of Perfect Ordinariness

## What Does It Mean to Be Quintessentially American?

Our statistical journey reveals profound insights about American identity:

### The Paradox of Exceptional Ordinariness

**Van Buren County, Michigan** and its fellow Goldilocks Zone counties achieve something remarkable: they are extraordinary in their ordinariness. Being perfectly average across 40+ variables requires a complex balancing act that most places cannot achieve.

### The Geographic Soul

The concentration of average counties in the Midwest suggests these regions serve as America's demographic anchor—places where various historical, economic, and cultural forces have combined to create representative communities.

### The Evolution of "Normal"

Our temporal analysis reveals that American averageness is not static. The counties that represented typical America in 2010 are not necessarily the same ones that do so today, reflecting the dynamic nature of our national identity.

## Policy Implications: The Value of the Goldilocks Zone

```{r policy-implications, fig.height=6}
# Analyze outcomes by averageness
outcomes_by_averageness <- county_full_analysis %>%
  mutate(
    averageness_decile = ntile(desc(mahal_distance), 10),
    averageness_category = case_when(
      averageness_decile <= 2 ~ "Most Average (Top 20%)",
      averageness_decile <= 4 ~ "Above Average (20-40%)",
      averageness_decile <= 6 ~ "Moderately Average (40-60%)",
      averageness_decile <= 8 ~ "Below Average (60-80%)",
      TRUE ~ "Least Average (Bottom 20%)"
    )
  ) %>%
  group_by(averageness_category) %>%
  summarise(
    n_counties = n(),
    mean_housing_stability = mean(housing_stability, na.rm = TRUE),
    mean_population_stability = mean(stability_proxy, na.rm = TRUE),
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    averageness_category = factor(averageness_category, 
                                  levels = c("Most Average (Top 20%)", "Above Average (20-40%)", 
                                           "Moderately Average (40-60%)", "Below Average (60-80%)", 
                                           "Least Average (Bottom 20%)"))
  )

# Create outcomes visualization
outcomes_plot <- outcomes_by_averageness %>%
  pivot_longer(cols = c(mean_housing_stability, mean_population_stability, mean_youth_ratio),
               names_to = "outcome", values_to = "value") %>%
  mutate(
    outcome_label = case_when(
      outcome == "mean_housing_stability" ~ "Housing Stability\n(% Homeownership)",
      outcome == "mean_population_stability" ~ "Population Stability\n(Lower Migration)",
      outcome == "mean_youth_ratio" ~ "Population Growth\nProxy (Youth Ratio)"
    )
  ) %>%
  ggplot(aes(x = averageness_category, y = value, fill = averageness_category)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~outcome_label, scales = "free_y", ncol = 3) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  labs(
    title = "Outcomes by Demographic Averageness",
    subtitle = "Do the most average counties show different stability and growth patterns?",
    x = "Averageness Category",
    y = "Outcome Value",
    caption = "Based on analysis of 3,000+ U.S. counties"
  ) +
  theme_goldilocks +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    strip.text = element_text(face = "bold", size = 10)
  )

outcomes_plot
```

### Strategic Value of Average Places

1. **Market Research Goldmine**: These counties offer ideal testing grounds for products and services targeting mainstream America

2. **Political Bellwethers**: Places achieving demographic balance may provide early signals of national political trends

3. **Policy Pilot Programs**: Testing new policies in representative communities before national rollout

4. **Housing Market Stability**: The correlation between averageness and homeownership suggests these places foster residential stability

# Peer Review & Critical Assessment

## Methodological Strengths and Limitations

**Strengths:**
- Multi-metric approach reveals different dimensions of averageness
- Temporal analysis shows evolution of American demographic center
- Comprehensive variable set captures multiple aspects of community life
- Sensitivity analysis tests robustness of findings

**Limitations:**
- County-level analysis masks within-county variation
- Mahalanobis distance assumes multivariate normality
- Some outcome measures are proxies rather than direct measurements
- 2018-2022 data may not capture post-pandemic shifts

## Alternative Interpretations

1. **Geographic bias**: Midwest dominance might reflect historical settlement patterns rather than inherent "averageness"
2. **Scale effects**: County boundaries vary dramatically in size and population
3. **Selection bias**: Available Census variables may not capture all relevant dimensions of community life

## Future Research Directions

1. **Causal mechanisms**: What drives certain places toward comprehensive averageness?
2. **Social capital**: How does demographic balance relate to community cohesion?
3. **Economic resilience**: Do average places weather economic shocks differently?
4. **Longitudinal outcomes**: Do children growing up in average places have different life trajectories?

# Conclusion: The Extraordinary Nature of the Ordinary

Van Buren County, Michigan, and its fellow travelers in the Goldilocks Zone represent something increasingly rare in American society: **comprehensive balance**. In an era of growing polarization and regional divergence, these places achieve what most cannot—they embody the statistical center of American life across dozens of dimensions simultaneously.

## The Deeper Truth

This whimsical investigation reveals a profound insight: **in a country of extremes, the middle ground itself has become extraordinary**. The counties achieving perfect demographic ordinariness are not boring—they are exceptional in their ability to balance the complex tensions that define American identity.

## The Value of the Center

As America grapples with questions of identity, representation, and belonging, these statistically average places offer something invaluable: a glimpse of what comprehensive American representation actually looks like when measured rigorously across multiple dimensions.

Perhaps most importantly, they remind us that underneath our differences lies a shared demographic reality—a statistical soul of America that persists even as our politics and culture seem to fragment.

In seeking the most ordinary places, we have discovered something extraordinary about what it means to be American in the 21st century: **true representativeness is an achievement, not an accident**.

---

## Technical Appendix

**Data Sources**: 2018-2022 American Community Survey 5-year estimates, 2010 ACS 5-year estimates  
**Geographic Coverage**: 3,000+ U.S. counties  
**Variables Analyzed**: 40+ demographic, economic, and social indicators  
**Statistical Methods**: Mahalanobis distance, Euclidean distance, median absolute deviation, quartile spread analysis  
**Software**: R with tidycensus, sf, tidyverse, and specialized visualization packages  

*Analysis conducted using multiple averageness metrics to ensure robustness of findings. All code and data available in accompanying .R script.*

---

*🤖 Generated with [Claude Code](https://claude.ai/code)*

*Co-Authored-By: Claude <noreply@anthropic.com>*